A free/open-source Kazakh-Tatar machine translation system

### 2013-04-22 paper/paper.tex
some updates

\usepackage{polyglossia}
\setdefaultlanguage[variant=australian]{english}
\usepackage{latexsym}
\usepackage[small,bf]{caption}
\usepackage{xltxtra}

\usepackage{fontspec}
\defaultfontfeatures{PunctuationSpace=3,Scale=MatchLowercase,Mapping=tex-text}
\newfontfeature{IPA}{+mgrk}
%\setromanfont[IPA]{FreeSerif}
\setromanfont[Scale=0.9]{Times New Roman}
\newfontfamily\qipa[IPA,Scale=MatchLowercase]{FreeSerif}
\newfontfamily\qipb[IPA,Scale=MatchLowercase]{Junicode}
\setmonofont[Scale=0.7]{DejaVu Sans Mono}
\newfontfamily\smallertt[Scale=0.55]{DejaVu Sans Mono}

@@

The bilingual lexicon currently contains 9,269 stem-to-stem correspondences and was built mostly by hand (i.e., by translating Kazakh stems unrecognized by the morphological analyzer into Tatar).  Some toponyms and other proper names were translated semi-automatically by looking up links in Wikipedia; also, some Russian loanwords common to both languages (such as автомобиль, гонорар, etc.) were added to the bilingual dictionary automatically by taking the intersection of Russian and Kazakh wordlists.

@@

%FIXME: JNW needs to copy-edit this, and we should probably provide a concrete example to make it clearer too
The output of the morphological analyzer is highly ambigious. Both Kazakh and Tatar have a lot of affixes which join the
verb stem. Some of the grammars label them all gerunds, or participles, assuming that there are no finite forms.

@@

\begin{table}[htbp]
	\centering
	\subfloat[Coverage of Kazakh transducer over Kazakh corpora]{
		\begin{tabular}{llrrr}
			\toprule
			Corpus          & Tokens & Coverage & stdev\\
			\midrule
			RFERL 2010      & 3.2M   & 90.26\% & 0.24\% \\
			RFERL 2012      & 2.9M   & 89.81\% & 0.59\% \\
			Wikipedia       & 1.2M   & 80.56\% & 5.44\% \\
			\bottomrule
		\end{tabular}
	}\\
	\subfloat[Coverage of Tatar transducer over Tatar corpora]{
		\begin{tabular}{llrrr}
			\toprule
			Corpus          & Tokens & Coverage & stdev\\
			\midrule
			RFERL 2007-2012 & 1.2M   & 82.56\%  & 3.24\% \\
			Wikipedia       & 128K   & 82.23\%  & 1.58\% \\
			New Testament   & 137K   & 92.54\%  & 1.31\% \\
			\bottomrule
		\end{tabular}
	}
	\caption{Naïve vocabulary coverage of the stand-alone transducers.}
	\label{table:coverage}
\end{table}

%FIXME: replace old text
As shown in Table~\ref{table:coverage}, the coverage is still far too low to be of use as a 
general broad-domain MT system, but we hope that it shows that a good proportion of 
the morphology of both languages is in place.

\begin{table}[htbp]
	\centering
	\begin{tabular}{llrrr}
		\toprule
		Corpus          & Tokens & Coverage & stdev\\
		\midrule
		RFERL 2010      & 3.2M   & 90.19\% & 0.23\% \\
		RFERL 2012      & 2.9M   & 89.74\% & 0.59\% \\
		Wikipedia       & 1.2M   & 80.75\% & 5.23\% \\
		\bottomrule
	\end{tabular}
	\caption{Naïve trimmed coverage of the Kazakh-Tatar system}
	\label{table:trimmedcoverage}
\end{table}

@@

To measure the performance of the translator we used the Word Error Rate metiric -- an edit metric based on the
Levenshtein distance \citep{levenshtein/1966}.

We had two small Kazakh corporas alongside with their postedited translations into Tatar to measure the WER.
The first one (2457 words total) was a concatenation of an article from Radio Free Europe / Radio Liberty's
Kazakh-language service, an article from Wikipedia, and a simple story used for pedagogical purposes in a workshop
on MT for the languages of Russia. In addition to postediting the translation, this corpus was run through the
morphological transducer and its output was manually disambiguated. All the stems in these texts had been added to
the system, and all the rules (Constraint Grammar,lexical selection and transfer) were based on this corpus.
Hence we can call it the development corpus. It presents an upper bound on the current performance
of the system. Table~\ref{table:wer-development} presents the WER for it. % http://pastebin.com/TaaWxJzN

The second corpus -- a collection of only Radio Free Europe / Radio Liberty's
Kazakh-language service articles -- was used solely for evaluation. It had a similar size -- 2862 words.
The WER results for it can be seen in the table~\ref{table:wer-development}. % http://pastebin.com/7tzGCgMX.

@@

\begin{table}[htbp]
  \begin{center}
  \begin{tabular}{ccrrr}
  \toprule
   Corpus                 & Direction           & Tokens  & Unknown & WER  \\
  \midrule
  Development corpus      & kaz$\rightarrow$tat & 2457    & 2       & 15.19\% \\
  \bottomrule
  \end{tabular}
    \caption{Word error rate over the development corpus.}
 %\vspace{-2.5em}
    \label{table:wer-development}
  \end{center}
\end{table}

\begin{table}[htbp]
  \begin{center}
  \begin{tabular}{ccrrr}
  \toprule
   Corpus                 & Direction           & Tokens  & Unknown & WER  \\
  \midrule
  Evaluation corpus       & kaz$\rightarrow$tat & 2862    & 43      & 36.57\% \\
  \bottomrule
  \end{tabular}
    \caption{Word error rate over the evaluation corpus.}
 %\vspace{-2.5em}
    \label{table:wer-evaluation}
  \end{center}
\end{table}

>>>>

\usepackage{polyglossia}
\setdefaultlanguage[variant=australian]{english}
\usepackage{latexsym}
%\usepackage[small,bf]{caption}
\usepackage[bf]{caption}
\usepackage{xltxtra}
\usepackage{times}
\usepackage{fontspec}
\defaultfontfeatures{PunctuationSpace=3,Scale=MatchLowercase,Mapping=tex-text}
\newfontfeature{IPA}{+mgrk}
%\setromanfont[IPA]{FreeSerif}
\setromanfont[Scale=0.9]{Times New Roman}
\newfontfamily\qipa[IPA,Scale=MatchLowercase]{FreeSerif}
\newfontfamily\qipb[IPA,Scale=MatchLowercase]{Junicode}
\setmonofont[Scale=0.7]{DejaVu Sans Mono}
\newfontfamily\smallertt[Scale=0.55]{DejaVu Sans Mono}

@@

The bilingual lexicon currently contains 9,269 stem-to-stem correspondences and was built mostly by hand (i.e., by translating Kazakh stems unrecognized by the morphological analyser into Tatar).  Some toponyms and other proper names were translated semi-automatically by looking up links in Wikipedia; also, some Russian loanwords common to both languages (such as автомобиль, гонорар, etc.) were added to the bilingual dictionary automatically by taking the intersection of Russian and Kazakh wordlists.

@@

%FIXME: JNW needs to copy-edit this, and we should probably provide a concrete example to make it clearer too
The output of the morphological analyser is highly ambigious. Both Kazakh and Tatar have a lot of affixes which join the
verb stem. Some of the grammars label them all gerunds, or participles, assuming that there are no finite forms.

@@

%%
%%\begin{table}[htbp]
%%	\centering
%%	\subfloat[Coverage of Kazakh transducer over Kazakh corpora]{
%%		\begin{tabular}{llrrr}
%%			\toprule
%%			Corpus          & Tokens & Coverage & stdev\\
%%			\midrule
%%			RFERL 2010      & 3.2M   & 90.26\% & 0.24\% \\
%%			RFERL 2012      & 2.9M   & 89.81\% & 0.59\% \\
%%			Wikipedia       & 1.2M   & 80.56\% & 5.44\% \\
%%			\bottomrule
%%		\end{tabular}
%%	}\\
%%	\subfloat[Coverage of Tatar transducer over Tatar corpora]{
%%		\begin{tabular}{llrrr}
%%			\toprule
%%			Corpus          & Tokens & Coverage & stdev\\
%%			\midrule
%%			RFERL 2007-2012 & 1.2M   & 82.56\%  & 3.24\% \\
%%			Wikipedia       & 128K   & 82.23\%  & 1.58\% \\
%%			New Testament   & 137K   & 92.54\%  & 1.31\% \\
%%			\bottomrule
%%		\end{tabular}
%%	}
%%	\caption{Naïve vocabulary coverage of the stand-alone transducers.}
%%	\label{table:coverage}
%%\end{table}

%FIXME: replace old text
As can be seen in Table~\ref{table:coverage}, the coverage for the news corpora approaches
that of a broad-coverage MT system, with one word in ten unknown. These unknown words are 
principally proper nouns. For the Wikipedia corpus, the coverage is substantially worse. This is 
due to \textbf{FIXME: why?}

\begin{table}
	\centering
	\begin{tabular}{llrrr}
		\toprule
		Corpus          & Tokens & Coverage & stdev\\
		\midrule
		RFERL 2010      & 3.2M   & 90.19\% & $\pm$ 0.23\% \\
		RFERL 2012      & 2.9M   & 89.74\% & $\pm$ 0.59\% \\
		Wikipedia       & 1.2M   & 80.75\% & $\pm$ 5.23\% \\
		\bottomrule
	\end{tabular}
	\caption{Naïve trimmed coverage of the Kazakh-Tatar system}
	\label{table:coverage}
\end{table}

@@

To measure the performance of the translator we used the Word Error Rate metiric --- an edit metric based on the
Levenshtein distance \citep{levenshtein/1966}.

We had two small Kazakh corporas alongside with their postedited translations into Tatar to measure the WER.
The first one (2457 words total) was a concatenation of an article from Radio Free Europe / Radio Liberty's
Kazakh-language service, an article from Wikipedia, and a simple story used for pedagogical purposes in a workshop
on MT for the languages of Russia. In addition to postediting the translation, this corpus was run through the
morphological transducer and its output was manually disambiguated. All the stems in these texts had been added to
the system, and all the rules (Constraint Grammar, lexical selection and transfer) were based on this corpus.
Hence we can call it the development corpus. It presents an upper bound on the current performance
of the system. Table~\ref{table:wer-development} presents the WER for it. % http://pastebin.com/TaaWxJzN

The second corpus --- a collection of only Radio Free Europe / Radio Liberty's
Kazakh-language service articles --- was used solely for evaluation. It had a similar size --- 2862 words.
The WER results for it can be seen in the table~\ref{table:wer-development}. % http://pastebin.com/7tzGCgMX.

@@

\begin{table}
  \begin{center}
  \begin{tabular}{ccrrr}
  \toprule
   Corpus                 & Direction           & Tokens  & OOV & WER (\%) \\
  \midrule
  \texttt{devel} & kaz$\rightarrow$tat & 2457    & 2       & 15.19 \\
  \bottomrule
  \texttt{test} & kaz$\rightarrow$tat & 2862    & 43      & 36.57 \\
  \end{tabular}
    \caption{Word error rate over two corpora. The \texttt{devel} `development' corpus was used during development 
       to identify possible disambiguation, lexical-selection and transfer rules. The \texttt{test} corpus
       was not seen during development. The OOV column gives the number of out-of-vocabulary (unknown) words.}
 %\vspace{-2.5em}
    \label{table:wer-development}
  \end{center}
\end{table}
